---
title: "Live Coding - Data Science Lifecycle"
format: html
editor: visual
execute:
  echo: true
  warning: false
editor_options: 
  chunk_output_type: console
---

```{r}
library(gapminder) # R package that contains the data
library(dplyr)
library(ggplot2)
library(knitr)
```

# Data import

Data can be imported from many different sources. In this exercise, we import data from:

1.  an R Package that is loaded via the `library()` function.

## Gapminder data

The data is an [excerpt of the Gapminder data](https://www.gapminder.org/data/documentation/) on life expectancy, GDP per capita, and population by country. Compiled for the purpose of teaching with data until 2007.

```{r}

```

# Data tidying

Once you imported the data, the next step in any data science lifecycle is the tidying of data. It's about bringing your data into a consistent structure that let's you focus on the analysis. Depending on how messy the data structure is, this can be tedious task and take up a great part of the data science lifecycle process.

## Gapminder data

# Data transformation

Once your data is tidy, a common first step is to transform it. This includes:

1.  narrowing in on what interests you (e.g. all people without toilets in one district, or all data from last year)

2.  creating new variables from existing (like toilet density, as in number of people per toilet)

3.  calculating summary statistics (like counts or the mean)

## Gapminder data

**Goal:** Calculate the median life expectancy at birth by continent for 2007.

-   Step 1: Filter the data to only include the year 2007

```{r}
gapminder_yr_2007 <- gapminder |> 
  filter(year == 2007)
```

-   Step 2: Group the data by continent and calculate the median life expectancy

```{r}

gapminder_summary_continent <- gapminder_yr_2007  |>
  group_by(continent) |> 
  summarise(lifeExp = median(lifeExp))
```

# Data visualisation (and tables)

## Gapminder data

Africa has the lowest life expectancy and greatest spread in life expectancy globally.

-   Use a boxplot with jittered points to visualize the median life expectancy by continent in 2007.

```{r}
ggplot(data = gapminder_yr_2007, 
       mapping = aes(x = continent, 
                     y = lifeExp)) +
  geom_boxplot() +
  geom_jitter(width = 0.1, alpha = 1/4, size = 3) +
  labs(x = NULL,
       y = "life expectancy") +
  theme_minimal(base_size = 14) 

```

-   Use a function to display the summary table of median life expectancy by continent.

```{r}

```

# Data communication

Your rendered html file is a great data communication product. You will learn a bit later in the course how you could make it accessible to others as a website, but you could also just download and share the html file with a collaborator or render the document to 'pdf' or 'docx' formats.

# Stage, Commit & Push to GitHub

This final section introduces the workflow you need to follow at the end of each live coding exercise, homework assignment, learning reflection. It ensures that all the changes you have made to files in this repository (folder) are synced with the version that is stored on GitHub. The steps will be described at the end of each of the files.

1.  Open the Git pane in RStudio. It's in the top right corner in a separate tab.

2.  **Stage** your changes by checking appropriate box next to all files (if you select one file with your mouse, you can then highlight them all with Ctrl + A on your keyboard and check all boxes)\$

3.  Write a meaningful commit message (e.g. "Completed homework assignment 1b) in the **Commit message** box

4.  Click **Commit**. Note that every commit needs to have a commit message associated with it.

5.  Go to your course-material GitHub repo and refresh the page. You should see your commit message next to the updated files.
